---
title: "chapter6.Rmd"
author: "Faranak Halali"
output: html_document
---

# Week 6: Analysis of longitudinal data
Reading data
```{r}
library(dplyr)
BPRSL <- read.csv("~/IODS-project/Data/BPRSL.csv")
BPRSL <- select(BPRSL, -X)
RATSL <- read.csv("~/IODS-project/Data/RATSL.csv")
RATSL <- select(RATSL, -X)
str(BPRSL)
str(RATSL)
```

'treatment' and 'subject' variables from the BPRSL data and 'ID' and 'Group' from the RATSL data should be converted into factor variables. Let's do this!
```{r}
library(dplyr)
library(tidyr)
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
glimpse(BPRSL)
glimpse(RATSL)
```

**Analyses of chapter 8 on the RATSL data**
```{r}
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

**Interpretation**: The plot shows that almost all rats in three groups of diet gain weight during the 8-week period, although the slope of weight gain is lower in group 1 compared to G2 and G3. Rats in groups 2 & 3 have higher baseline weights than the rats in G1. In G1, the final weight of the rats are similar to each other (except the one rat which has the lowest baseline and also the lowest final weight). Rats in G3 have also the same situation. However, in G2 there is one rat which is significantly different from the others in this group: it has much higher baseline and final weight than the others. Also there is one rat in this group (solid black line) which despite its high baseline weight it does not gain weight differently than the other 2 rats and they all reach a similar final weight. Overall, I am not sure whether the *tracking* effect exists in this study because the rats which have higher baseline weigth tend to have higher weight also throughout the study whereas their final weights are quite similar (there are exceptions of course).

Standardization of the RATSL data
```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)
```

Plot the standardized RATSL data to see the individual variabilities more clearly
```{r}
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized Weight")
```

Summary data of the RATSL data
```{r}
n <- RATSL$Time %>% unique() %>% length()
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)
```

Now, RATSS data which is the summary data of RATSL has 33 observations and 4 variables (compared to  RATSL data which had 176 observations and 6 variables). RATSS data shows the mean and standard error of weight for each group in each Time point, e.g., mean weight of group 1 in day 1, mean weight of G1 in day 8, so on. 

Summary plot of the RATSS data
```{r}
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

From the plot, it is clear that the average weight of all three groups has increased during the study, but the rate of increase is much slower in group 1 than the other 2 groups and this could be due to the lower baseline weight of this group. Group 3 which has the highest baseline average weight has also the highest final average weight. 

Boxplot for the RATSL data
```{r}
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
glimpse(RATSL8S)
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-8")
```

This boxplot confirms the results of the previous summary plot plus that there is one outlier observation in each group. The outlier of the G2 is above the quartile 3 whereas the outliers of the G1 and G3 are both below the quartile 1.

Filtering the outlier and drawing the plot again
```{r}
RATSL8S1 <- RATSL8S %>%
  filter(mean < 550)
RATSL8S2 <- RATSL8S1 %>%
  filter(mean > 250)
ggplot(RATSL8S2, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-8")
```

I defined the two ends of the threshold for filtering the outliers from G1 and G2 (mean > 250 & mean < 550), but could not excluded the outlier observation from G3. 

ANOVA analysis on RATSL8S2 data
```{r}
anova <- aov(RATSL8S2$mean ~ RATSL8S2$Group)
summary(anova)
```

Because our aim is to compare the difference between three independent groups we had to use ANOVA. The results of the ANOVA indicate that there is a significant main effect of group on rats weight gain (F(2,11)=611.4, p<.001), i.e., there is a significant difference (p<0.001) in mean weight gain between the three groups of rats. However, we do not know which groups differ from each other and for that information we have to perform Post hoc tests. Let's do this!

Performing Post hoc Tukey test and its plot
```{r}
tukey <- TukeyHSD(anova)
tukey
plot(tukey)
```
The output contains the mean difference between each of group pairs, lower end and upper end of the 95% confidence interval and the adjusted p-values. The results of the Tukey test show that all three group pairs are significantly different from each other (G1&G2, G1&G3, G2&G3). 


**Analyses of chapter 9 on the BPRS data**
```{r}
dim(BPRSL)
ggplot(BPRSL, aes(x = week, y = bprs, group = interaction(subject, treatment))) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "week", breaks = seq(0, 8, 4)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
```

This plot shows how the bprs scores of the subjects of each treatment group have changed during the 8 weeks of the study. 

Fitting a linear model
```{r}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
```

Results of the linear model indicate that treatment 2 is not significantly different from the treatment 1 regarding bprs. Also, it shows that week has a significant main effect on bprs. This model explains about 18% of the variance in bprs. 


Random intercept model
```{r}
library(lme4)
library("lmerTest")
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
plot(BPRS_ref)
anova(BPRS_ref)
```

The random intercept model allows the linear regression fit for each subject to differ in intercept from other subjects. In this model we can see that there is quite a variation in the random facotr *subject* (std. Dev=6.885). Comparing this model with the previous linear regression model shows that the *Estimated regression parameters* value do not change with including random effects whereas standard errors are affected (smaller) in the random intercept model. The random intercept model indicates that week has a main effect on bprs also when subject is included as a random factor. The fitted vs. residuals plot does not show a certain pattern and the data points are symmetrically dispersed around the y=0 line (linearity of the model) and there is no obvious outlier.


Fitting a random intercept and random slope model
```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref1, BPRS_ref)
plot(BPRS_ref1)
```

The results for the fixed effects are very similar to those of the random intercept model, but the likelihood ratio test for the random intercept model versus the random intercept and slope model gives a chi-squared statistic of 7.27 and the associated p-value of 0.026. Since the chi-squared value is rather small we can say that compared to the random intercept model the random intercept and slope model provides a better fit for these data. Again, the fitted vs. residuals plot does not show a certain pattern and the data points are symmetrically dispersed around the y=0 line (linearity of the model) and there is no obvious outlier.


Fitting a random intercept and random slope model with the interaction
```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
anova(BPRS_ref2, BPRS_ref1)
plot(BPRS_ref2)
```

In this model, we also consider the interaction between week and treatment, i.e., to study whether some subjects of a specific treatment group in certain weeks have different bprs profile. From the output we can say that the interaction between week and treatment is not significant (t-value < 2, p-value > 0.05). Also, comparing the chi-squared value and its related p-value we can say that there is no significant difference between this interaction model and the previous random intercept and random slope model and thus the interaction model does not add any new information. As for the previous two models, the fitted vs. residuals plot does not show a certain pattern and the data points are symmetrically dispersed around the y=0 line (linearity of the model) and there is no obvious outlier.


Drawing the plot of BPRSL
```{r}
ggplot(BPRSL, aes(x = week, y = bprs, group = interaction(subject, treatment))) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "week", breaks = seq(0, 8, 4)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
```

This is the same plot as we saw before and it shows the change in bprs in subjects of the two treatment groups throughout the 8 weeks of the study.

Creating a vector of the fitted values and adding a new column
```{r}
Fitted <- fitted(BPRS_ref2)
BPRSL <- BPRSL %>%
  mutate(Fitted)
```

Drawing the plot of BPRSL with fitted values
```{r}
ggplot(BPRSL, aes(x = week, y = Fitted, group = interaction(subject, treatment))) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "week", breaks = seq(0, 8, 4)) +
  scale_y_continuous(name = "Fitted bprs") +
  theme(legend.position = "top")
```

This plot, which is based on fitted (predicted) bprs values, suggests that the change in bprs scores of subjects in two treatment groups would be like as in the graph. However, we saw that this linear change does not happen in real setting and there is kind of a fluctutation in bprs scores throughout the 8 weeks of the study. 